Introductory Physics Labs: A Tale of Two Transformations

### 2020-08-25 manuscript/onlineLabPaper.tex
Recovered section 3, which was inadvertently deleted.
Updated draft with Instructional context and pedagogical changes.  Ready for meeting.  We aren't massively over on word count... $ texcount onlineLabPaper.tex File: onlineLabPaper.tex Encoding: ascii Words in text: 3371 Words in headers: 50 Words outside text (captions, etc.): 73 Number of headers: 12 Number of floats/tables/figures: 5 Number of math inlines: 2 Number of math displayed: 0 Subcounts:   text+headers+captions (#headers/#floats/#inlines/#displayed)   88+9+0 (2/0/0/0) _top_   378+1+0 (1/0/0/0) Section: Introduction   366+2+39 (1

\author{Steven F.\ Wolf} \affiliation{East Carolina University Department of Physics 1000 E.\
  5\textsuperscript{th} street, Greenville, NC 27858 USA} \author{Mark W.\ Sprague}
\affiliation{East Carolina University Department of Physics 1000 E.\ 5\textsuperscript{th}
  street, Greenville, NC 27858 USA}

@@

\section{Instructional Context}
\textcolor{blue}{SFW mostly writes this.}

We should also talk about the specific context of our labs:
\begin{itemize}
  \item Discuss institutional profile.
  \item Discuss student population in labs/lecture courses served.
  \item Compare course demographics with university demographics.
\end{itemize}

>>>>

% \author{Steven F.\ Wolf} \affiliation{East Carolina University Department of Physics 1000 E.\
% 5\textsuperscript{th} street, Greenville, NC 27858 USA} \author{Mark W.\ Sprague}
% \affiliation{East Carolina University Department of Physics 1000 E.\ 5\textsuperscript{th}
% street, Greenville, NC 27858 USA}

\author{First Author} \affiliation{Masked Institution} \author{Second Author}
\affiliation{Masked Institution}

@@

\section{Instructional Context}
\textcolor{red}{This section needs to have the institution masked before we send it in.}

East Carolina University is a regional master's university in a rural, economically depressed
part of our state.  The university mission is focused on regional transformation and service to
this region.  The total student population in Fall 2019 was 23k.  Our physics labs (Physics 1
lab and Physics 2 lab) are each 1 credit courses that meet for 2 hours per week. These courses
serve both of our Calculus-based and Algebra-based physics lecture courses.  Most $(\sim 75\%)$
of the students who take our lab course are in (or have taken) the Algebra-based physics
lecture course. It also fulfills the general education requirement for science lab courses at
our university.  Table \ref{tab:gender} compares the gender of students in these lab courses
compared to the university population.  We note that the gender distribution of students in the
Physics 2 lab generally matches the university population, while the Physics 1 lab skews more
heavily male than the university population.  We also compare the race of students in these
contexts in Table \textcolor{red}{not ready yet}\ldots

Students who take these labs are most often science majors, especially in the health sciences
(e.g., Biology, Exercise Science), but this lab is also taken by our majors.  In Fall 2019 we
had 409 students in the Physics 1 lab, and in Spring 2020 (at Census Day, before the pandemic
changed enrollments) we had 256 students in the Physics 2 lab.  In both labs, our enrollment is
capped at 22 students, due to the constraints of our laboratory classroom.

Labs are supervised by MWS, but each section is run by Graduate TAs (GTAs). The transformed
curriculum was jointly written by both of the authors.  TA training was greatly enhanced when
the new lab curriculum was put in place.  SFW and MWS, along with colleagues in Biology and
Chemistry, run a training for all GTAs in Biology, Chemistry, Geology, and Physics as these
disciplines are all using the same curricular format for their labs.  MWS runs a weekly prep
meeting with the TAs, and SFW and the research team have supervised various aspects of the
transformation especially important for research such as curricular implementation \cite{AL's
  paper} and assessment practices \cite{Wolf2019}.

\begin{table}
  \centering
  \begin{tabular}{l|ll|l}
    \hline \hline							
    &	\multicolumn{2}{c|}{Course}\\
    Gender	&	Physics 1	&	Physics 2	&	University	\\ \hline
    Female	&	41.8\%	&	60.9\%	&	57.1\%	\\
    Male	&	58.2\%	&	39.1\%	&	42.9\%	\\
    \hline \hline							
  \end{tabular}
  \caption{Gender breakdown of students in the Fall 2019 of the Physics 1 lab, the Spring 2020
    Physics 1 lab, and the undergraduate poplulation of the university at large in Fall 2019.
    Note: The university only collected binary gender data }
  \label{tab:gender}
\end{table}

\section{Transformation \#1: Argument Driven Inquiry}
Argument Driven Inquiry (ADI) is a pedagogical method that was developed to make science labs
in school better match what actually transpires in authentic research settings
\cite{Walker2011,Sampson2011,Walker2016}.  In traditional labs, we often ask students to answer
questions, or measure quantities, that are well-known using procedures that have been refined
over time.  For example, in physics we often have students measure $g$ using a well-known
method (such as a spark timer).  In ADI labs, we give students a question to answer (e.g.,
``Does the force a fan exerts on a cart depend on the mass of the cart?''), allow them to
design a procedure for an experiment that will answer that question, collect and analyze their
data, and make claims based on their data.  We accomplish this by having students engage in a
three-part cycle:
\begin{itemize}
  \item Pre-Lab - Introducing the context
  \item Proposal Development and Data Collection
  \item Argumentation and Peer Review
\end{itemize}
In the sections below, we will comment on each part of the cycle, focusing on both students and
GTAs.

\subsection{Week 1: Pre-Lab}
Part of the goal of any laboratory is to help students develop a familiarity with certain tools
and equipment common to the lab.  The purpose of the pre-lab reading and activity is to give
students familiarity with the equipment they will be using.  For the investigation that I
described previously, ``Does the force a fan exerts on a cart depend on the mass of the cart?''
we need students to become familiar with a few things, such as what we mean by a cart, and give
them some tools that they need to be able to answer this question.  As a part of the pre-lab
reading, students are given reading about linear regression---both how to run it in Excel, and
how to interpret the output, and how we can estimate instantaneous velocity given position
vs. time data. For this experiment, we give them a cart (leaving the fan off for now) and have
them measure the acceleration of that cart down a ramp using an ultrasonic motion detector to
collect position vs.\ time data for this motion.  Students work in pairs to do this work, and
turn in a brief summary of this work on the course management system.  For this investigation,
it is the first time that they create graphs, so we set norms for graphing such as helping them
decide which variable to put on the x-axis and the y-axis as well as including units.  Pre-lab
assignments are designed to be graded in less than 1 minute/paper based on a rubric we provide
to both students and GTAs.

\subsection{Week 2: Proposal Development and Data Collection}
During this week, students work in groups of 4 to develop a proposal, get it approved by a GTA,
and then collect and analyze their data.  We have a format for proposal development that asks
students to link the scientific concept being studied---in this example case, Newton's Second
Law---to the data that they will collect.  Students also need to propose a plan for analyzing
their data and minimizing potential sources of error (such as friction).  Once students have an
approved proposal, they begin collecting and analyzing their data.  This may lead them to
refine or revise their method (which they are encouraged to do).  Grading for the GTAs this
week is minimal (full credit for completing the proposal).  However, this is the most critical
week's for GTA training.  GTAs have to be clear that the goal is to allow a diverse number of
ways of collecting/analyzing this data.  How much does the mass of the cart need to vary?  How
many different cart masses do the students need to take?  How should students deal with
friction? All of these are questions that we want the \textit{students} to answer.  And in some
sense, we will be disappointed if all of the students come up with the same answer.
Experimental methods certainly have an impact on experimental results.

\subsection{Week 3: Argumentation and Peer Review}
During this week, the students begin by sharing their results.  Students finish analyzing their
data during the prior week, and prepare a ``poster'' on a whiteboard.  One person from the
group stays at the whiteboard while the other group members (travelers) go to other groups.
The travelers learn what other groups did, and compare to their results in a structured
argumentation setting.  Groups spend about 3-5 minutes at a poster, and rotate so that they see
at least 3 other posters.  After the travelers return, critically, the groups spend time
discussing what they saw and reflecting on their own results.  What is exciting is that
sometimes, these interactions actually drive groups to change their claims \cite{Walker2019}.

Students write a shortened lab report consisting of three sections: An introduction to the
scientific concept, a description of the experimental method or procedure, and a discussion
section laying out their results, and the evidence and justification supporting their results. 
The authors have created different peer review calibration videos for the students to watch, so
that they can look for the same things that we are going to be grading them on when we get
their final reports.  Students complete their lab report and turn it in by the end of the day.
Then, students take part in double-confidential peer review.  We facilitate this online.  They
read and reply to two different lab reports.  Finally, they get their feedback and use this to
revise their own lab report.  At the end of this week, students turn in their revised,
peer-reviewed final lab report.  This is the heaviest week of GTA grading.  We provide a rubric
for lab report grading, and each lab report generally takes about 3-5 minutes to grade.
